---
title: "Spinal Cord Injury Challenge Exploratory Analysis"
author: "Boris Polanco"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goal

The goal is to predict neurologic status at 6-13 moths after traumatic spinal cord injury based on acute phase assessments one week after injury. We require to prognosticate long-term recovery after SCI by analyzing a subset of the data from the Sygen multicenter clinical trial. The data used to make the prediction of recovery at 6-13 months will be data available in the first month after spinal cord injury. One piece of data that is commonly used to make such predictions are the internacional Standards for Neurological Classification of SCI (INCSCI). This clinical outcome assessment evaluates motor and sensory impairment across the body and helpts localize the neurologic level of the SCI and grade the severity of that injury. Additional data includes demographic information, comorbidities, surgical interventions and medical complications.

## Data

The following files are provided and explained based on the description given at <https://asia-spinalinjury.org/>.

### File 1: metada.csv

Includes time-invariante features (including age, bmi, but also information on previous conditions). PID is the patient ID.

```{r, warning=FALSE,message=FALSE,echo=FALSE }
library(dplyr)
library(ggplot2)
library(readr)


metadata <- read_csv("G:/My Drive/Thesis PhD Boris Polanco/ASIA_CHALLENGE_SCI/ASIAChallenge_ShareFile/metadata.csv",col_names = TRUE)

metadata
# The available variables are
colnames(metadata)
```

------------------------------------------------------------------------

------------------------------------------------------------------------

## Variable Descriptions

| Variable | Type | Description | Values |
|----------------|----------------|----------------|------------------------|
| PID | Fixed | patient id | \- |
| age_category | Fixed | age at time of injury either above or below 65 years | \<65, \>65 |
| sexcd | Fixed | sex | 1, female \| 2, male |
| bmi_category | Fixed | bmi categorized | underweight, healthy, overweight |
| srdecc1 | Fixed | first spinal surgery: were spinal roots decompressed during surgery | 0, no \| 1, yes |
| surgcd1 | Fixed | first spinal surgery: was spinal surgery performed | 0, no \| 1, yes |
| spcsuc1 | Fixed | first spinal surgery: Are there more than one area of spinal column surgery | 0, no \| 1, yes |
| scdecc1 | Fixed | first spinal surgery: was the spinal cord decompressed | 0, no \| 1, yes |
| hemccd1 | Fixed | first spinal surgery: did significant changes in the hemodynamic or pulmonary function occur during the operation | 0, no \| 1, yes |
| mhpsyccd | Fixed | has the patient had a past history of psychiatric abnormalities | 0, no \| 1, yes \| 9, unknown |
| mhneurcd | Fixed | has the patient had a past history of neurological abnormalities | 0, no \| 1, yes \| 9, unknown |
| mhcardcd | Fixed | has the patient had a past history of cardiac abnormalities | 0, no \| 1, yes \| 9, unknown |
| mhmetacd | Fixed | has the patient had a past history of metabolic or endocrinologic abnormalities | 0, no \| 1, yes \| 9, unknown |
| tx1_r | Fixed | treatment group | D1, dose 1 \| D2, dose 2 \| P, placebo |

### File 2: train_features.csv

All features, including ISNCSCI and AIS grades, at the time points 1,4,8 and 16 weeks after injury. Each row corresponds a specific patient ID, columns are features at various time points as given in the column names

```{r , echo=FALSE,message=FALSE}
train_features <- read_csv("G:/My Drive/Thesis PhD Boris Polanco/ASIA_CHALLENGE_SCI/ASIAChallenge_ShareFile/train_features.csv",col_names = TRUE)

```

## AIS grade and motor scores

| Variable   | Time Points | Description  | Side | Values  |
|------------|------------|--------------|------|---------|
| ais        | 1,4,8,16   | Calculated AIS grade at baseline  | - | A, B, C, D, E |
| elbfll     | 1,4,8,16   | Motor score C5  | Left | 0,1,2,3,4,5 |
| wrextl     | 1,4,8,16   | Motor score C6  | Left | 0,1,2,3,4,5 |
| elbexl     | 1,4,8,16   | Motor score C7  | Left | 0,1,2,3,4,5 |
| finfll     | 1,4,8,16   | Motor score C8  | Left | 0,1,2,3,4,5 |
| finabl     | 1,4,8,16   | Motor score T1  | Left | 0,1,2,3,4,5 |
| hipfll     | 1,4,8,16   | Motor score L2  | Left | 0,1,2,3,4,5 |
| kneexl     | 1,4,8,16   | Motor score L3  | Left | 0,1,2,3,4,5 |
| ankdol     | 1,4,8,16   | Motor score L4  | Left | 0,1,2,3,4,5 |
| gretol     | 1,4,8,16   | Motor score L5  | Left | 0,1,2,3,4,5 |
| ankpll     | 1,4,8,16   | Motor score S1  | Left | 0,1,2,3,4,5 |
| elbflr     | 1,4,8,16   | Motor score C5  | Right | 0,1,2,3,4,5 |
| wrextr     | 1,4,8,16   | Motor score C6  | Right | 0,1,2,3,4,5 |
| elbexr     | 1,4,8,16   | Motor score C7  | Right | 0,1,2,3,4,5 |
| finflr     | 1,4,8,16   | Motor score C8  | Right | 0,1,2,3,4,5 |
| finabr     | 1,4,8,16   | Motor score T1  | Right | 0,1,2,3,4,5 |
| hipflr     | 1,4,8,16   | Motor score L2  | Right | 0,1,2,3,4,5 |
| kneetr     | 1,4,8,16   | Motor score L3  | Right | 0,1,2,3,4,5 |
| ankdor     | 1,4,8,16   | Motor score L4  | Right | 0,1,2,3,4,5 |
| gretor     | 1,4,8,16   | Motor score L5  | Right | 0,1,2,3,4,5 |
| ankplr     | 1,4,8,16   | Motor score S1  | Right | 0,1,2,3,4,5 |

## Light touch scores

| Variable   | Time Points | Description  | Side | Values  |
|------------|------------|--------------|------|---------|
| c2ltl      | 1,4,8,16   | Light touch C2  | Left | 0,1,2 |
| c3ltl      | 1,4,8,16   | Light touch C3  | Left | 0,1,2 |
| c4ltl      | 1,4,8,16   | Light touch C4  | Left | 0,1,2 |
| c5ltl      | 1,4,8,16   | Light touch C5  | Left | 0,1,2 |
| c6ltl      | 1,4,8,16   | Light touch C6  | Left | 0,1,2 |
| c7ltl      | 1,4,8,16   | Light touch C7  | Left | 0,1,2 |
| c8ltl      | 1,4,8,16   | Light touch C8  | Left | 0,1,2 |
| t1ltl      | 1,4,8,16   | Light touch T1  | Left | 0,1,2 |
| t2ltl      | 1,4,8,16   | Light touch T2  | Left | 0,1,2 |
| t3ltl      | 1,4,8,16   | Light touch T3  | Left | 0,1,2 |
| t4ltl      | 1,4,8,16   | Light touch T4  | Left | 0,1,2 |
| t5ltl      | 1,4,8,16   | Light touch T5  | Left | 0,1,2 |
| t6ltl      | 1,4,8,16   | Light touch T6  | Left | 0,1,2 |
| t7ltl      | 1,4,8,16   | Light touch T7  | Left | 0,1,2 |
| t8ltl      | 1,4,8,16   | Light touch T8  | Left | 0,1,2 |
| t9ltl      | 1,4,8,16   | Light touch T9  | Left | 0,1,2 |
| t10ltl     | 1,4,8,16   | Light touch T10  | Left | 0,1,2 |
| t11ltl     | 1,4,8,16   | Light touch T11  | Left | 0,1,2 |
| t12ltl     | 1,4,8,16   | Light touch T12  | Left | 0,1,2 |
| l1ltl      | 1,4,8,16   | Light touch L1  | Left | 0,1,2 |
| l2ltl      | 1,4,8,16   | Light touch L2  | Left | 0,1,2 |
| l3ltl      | 1,4,8,16   | Light touch L3  | Left | 0,1,2 |
| l4ltl      | 1,4,8,16   | Light touch L4  | Left | 0,1,2 |
| l5ltl      | 1,4,8,16   | Light touch L5  | Left | 0,1,2 |
| s1ltl      | 1,4,8,16   | Light touch S1  | Left | 0,1,2 |
| s2ltl      | 1,4,8,16   | Light touch S2  | Left | 0,1,2 |
| s3ltl      | 1,4,8,16   | Light touch S3  | Left | 0,1,2 |
| s45ltl     | 1,4,8,16   | Light touch S4/5  | Left | 0,1,2 |
| c2ltr      | 1,4,8,16   | Light touch C2  | Right | 0,1,2 |
| c3ltr      | 1,4,8,16   | Light touch C3  | Right | 0,1,2 |
| c4ltr      | 1,4,8,16   | Light touch C4  | Right | 0,1,2 |
| c5ltr      | 1,4,8,16   | Light touch C5  | Right | 0,1,2 |
| c6ltr      | 1,4,8,16   | Light touch C6  | Right | 0,1,2 |
| c7ltr      | 1,4,8,16   | Light touch C7  | Right | 0,1,2 |
| c8ltr      | 1,4,8,16   | Light touch C8  | Right | 0,1,2 |
| t1ltr      | 1,4,8,16   | Light touch T1  | Right | 0,1,2 |
| t2ltr      | 1,4,8,16   | Light touch T2  | Right | 0,1,2 |
| t3ltr      | 1,4,8,16   | Light touch T3  | Right | 0,1,2 |
| t4ltr      | 1,4,8,16   | Light touch T4  | Right | 0,1,2 |
| t5ltr      | 1,4,8,16   | Light touch T5  | Right | 0,1,2 |
| t6ltr      | 1,4,8,16   | Light touch T6  | Right | 0,1,2 |
| t7ltr      | 1,4,8,16   | Light touch T7  | Right | 0,1,2 |
| t8ltr      | 1,4,8,16   | Light touch T8  | Right | 0,1,2 |
| t9ltr      | 1,4,8,16   | Light touch T9  | Right | 0,1,2 |
| t10ltr     | 1,4,8,16   | Light touch T10  | Right | 0,1,2 |
| t11ltr     | 1,4,8,16   | Light touch T11  | Right | 0,1,2 |
| t12ltr     | 1,4,8,16   | Light touch T12  | Right | 0,1,2 |
| l1ltr      | 1,4,8,16   | Light touch L1  | Right | 0,1,2 |
| l2ltr      | 1,4,8,16   | Light touch L2  | Right | 0,1,2 |
| l3ltr      | 1,4,8,16   | Light touch L3  | Right | 0,1,2 |
| l4ltr      | 1,4,8,16   | Light touch L4  | Right | 0,1,2 |
| l5ltr      | 1,4,8,16   | Light touch L5  | Right | 0,1,2 |
| s1ltr      | 1,4,8,16   | Light touch S1  | Right | 0,1,2 |
| s2ltr      | 1,4,8,16   | Light touch S2  | Right | 0,1,2 |
| s3ltr      | 1,4,8,16   | Light touch S3  | Right | 0,1,2 |
| s45ltr     | 1,4,8,16   | Light touch S4/5  | Right | 0,1,2 |


## Prin Prick

| Variable   | Time Points | Description  | Side | Values  |
|------------|------------|--------------|------|---------|
| c2ppl      | 1,4,8,16   | Pin prick C2  | Left | 0,1,2 |
| c3ppl      | 1,4,8,16   | Pin prick C3  | Left | 0,1,2 |
| c4ppl      | 1,4,8,16   | Pin prick C4  | Left | 0,1,2 |
| c5ppl      | 1,4,8,16   | Pin prick C5  | Left | 0,1,2 |
| c6ppl      | 1,4,8,16   | Pin prick C6  | Left | 0,1,2 |
| c7ppl      | 1,4,8,16   | Pin prick C7  | Left | 0,1,2 |
| c8ppl      | 1,4,8,16   | Pin prick C8  | Left | 0,1,2 |
| t1ppl      | 1,4,8,16   | Pin prick T1  | Left | 0,1,2 |
| t2ppl      | 1,4,8,16   | Pin prick T2  | Left | 0,1,2 |
| t3ppl      | 1,4,8,16   | Pin prick T3  | Left | 0,1,2 |
| t4ppl      | 1,4,8,16   | Pin prick T4  | Left | 0,1,2 |
| t5ppl      | 1,4,8,16   | Pin prick T5  | Left | 0,1,2 |
| t6ppl      | 1,4,8,16   | Pin prick T6  | Left | 0,1,2 |
| t7ppl      | 1,4,8,16   | Pin prick T7  | Left | 0,1,2 |
| t8ppl      | 1,4,8,16   | Pin prick T8  | Left | 0,1,2 |
| t9ppl      | 1,4,8,16   | Pin prick T9  | Left | 0,1,2 |
| t10ppl     | 1,4,8,16   | Pin prick T10  | Left | 0,1,2 |
| t11ppl     | 1,4,8,16   | Pin prick T11  | Left | 0,1,2 |
| t12ppl     | 1,4,8,16   | Pin prick T12  | Left | 0,1,2 |
| l1ppl      | 1,4,8,16   | Pin prick L1  | Left | 0,1,2 |
| l2ppl      | 1,4,8,16   | Pin prick L2  | Left | 0,1,2 |
| l3ppl      | 1,4,8,16   | Pin prick L3  | Left | 0,1,2 |
| l4ppl      | 1,4,8,16   | Pin prick L4  | Left | 0,1,2 |
| l5ppl      | 1,4,8,16   | Pin prick L5  | Left | 0,1,2 |
| s1ppl      | 1,4,8,16   | Pin prick S1  | Left | 0,1,2 |
| s2ppl      | 1,4,8,16   | Pin prick S2  | Left | 0,1,2 |
| s3ppl      | 1,4,8,16   | Pin prick S3  | Left | 0,1,2 |
| s45ppl     | 1,4,8,16   | Pin prick S4/5  | Left | 0,1,2 |
| c2ppr      | 1,4,8,16   | Pin prick C2  | Right | 0,1,2 |
| c3ppr      | 1,4,8,16   | Pin prick C3  | Right | 0,1,2 |
| c4ppr      | 1,4,8,16   | Pin prick C4  | Right | 0,1,2 |
| c5ppr      | 1,4,8,16   | Pin prick C5  | Right | 0,1,2 |
| c6ppr      | 1,4,8,16   | Pin prick C6  | Right | 0,1,2 |
| c7ppr      | 1,4,8,16   | Pin prick C7  | Right | 0,1,2 |
| c8ppr      | 1,4,8,16   | Pin prick C8  | Right | 0,1,2 |
| t1ppr      | 1,4,8,16   | Pin prick T1  | Right | 0,1,2 |
| t2ppr      | 1,4,8,16   | Pin prick T2  | Right | 0,1,2 |
| t3ppr      | 1,4,8,16   | Pin prick T3  | Right | 0,1,2 |
| t4ppr      | 1,4,8,16   | Pin prick T4  | Right | 0,1,2 |
| t5ppr      | 1,4,8,16   | Pin prick T5  | Right | 0,1,2 |
| t6ppr      | 1,4,8,16   | Pin prick T6  | Right | 0,1,2 |
| t7ppr      | 1,4,8,16   | Pin prick T7  | Right | 0,1,2 |
| t8ppr      | 1,4,8,16   | Pin prick T8  | Right | 0,1,2 |
| t9ppr      | 1,4,8,16   | Pin prick T9  | Right | 0,1,2 |
| t10ppr     | 1,4,8,16   | Pin prick T10  | Right | 0,1,2 |
| t11ppr     | 1,4,8,16   | Pin prick T11  | Right | 0,1,2 |
| t12ppr     | 1,4,8,16   | Pin prick T12  | Right | 0,1,2 |
| l1ppr      | 1,4,8,16   | Pin prick L1  | Right | 0,1,2 |
| l2ppr      | 1,4,8,16   | Pin prick L2  | Right | 0,1,2 |
| l3ppr      | 1,4,8,16   | Pin prick L3  | Right | 0,1,2 |
| l4ppr      | 1,4,8,16   | Pin prick L4  | Right | 0,1,2 |
| l5ppr      | 1,4,8,16   | Pin prick L5  | Right | 0,1,2 |
| s1ppr      | 1,4,8,16   | Pin prick S1  | Right | 0,1,2 |
| s2ppr      | 1,4,8,16   | Pin prick S2  | Right | 0,1,2 |
| s3ppr      | 1,4,8,16   | Pin prick S3  | Right | 0,1,2 |
| s45ppr     | 1,4,8,16   | Pin prick S4/5  | Right | 0,1,2 |


### File 3: train_outcomes_MS.csv

ISNCSCI motor scores assessed at either 26 or 52 weeks (as indicated in column "time"). Rows indicate patient IDs.

```{r , echo=FALSE,message=FALSE}
train_outcomes_MS <- read_csv("G:/My Drive/Thesis PhD Boris Polanco/ASIA_CHALLENGE_SCI/ASIAChallenge_ShareFile/train_outcomes_MS.csv",col_names = TRUE)

```


| Variable   | Time Points | Description  | Side | Values  |
|------------|------------|--------------|------|---------|
| modben     | 26, 52     | Modified Benzel Classification | - | 1, no motor or sensory function is preserved in the segments S4 to S5 |
|            |           |                                |   | 2, sensory but not motor function is preserved in the sacral segments S4 to S5 |
|            |           |                                |   | 3, motor function is preserved below the neurological level and the majority of key muscles below the neurological level have a muscle grade less than three and they are unable to walk |
|            |           |                                |   | 4, unable to walk and some functional motor control below the level of injury that is significantly useful (for example assist in transfers) but that is not sufficient for independent walking |
|            |           |                                |   | 5, limited walking and motor function allows walking with assistance or unassisted but significant problems secondary to lack of endurance or fear of falling limit patient mobility |
|            |           |                                |   | 6, unlimited walking and ambulatory without assistance and without significant limitations other than one or both of the following (difficulties with micturition or slightly dyscoordinated gait) |
|            |           |                                |   | 7, neurologically intact with the exception of minimal deficits that cause no functional difficulties |
|            |           |                                |   | 9, not able to assess |
| elbfll     | 26, 52     | Motor score C5  | Left  | 0,1,2,3,4,5 |
| wrextl     | 26, 52     | Motor score C6  | Left  | 0,1,2,3,4,5 |
| elbexl     | 26, 52     | Motor score C7  | Left  | 0,1,2,3,4,5 |
| finfll     | 26, 52     | Motor score C8  | Left  | 0,1,2,3,4,5 |
| finabl     | 26, 52     | Motor score T1  | Left  | 0,1,2,3,4,5 |
| hipfll     | 26, 52     | Motor score L2  | Left  | 0,1,2,3,4,5 |
| kneexl     | 26, 52     | Motor score L3  | Left  | 0,1,2,3,4,5 |
| ankdol     | 26, 52     | Motor score L4  | Left  | 0,1,2,3,4,5 |
| gretol     | 26, 52     | Motor score L5  | Left  | 0,1,2,3,4,5 |
| ankpll     | 26, 52     | Motor score S1  | Left  | 0,1,2,3,4,5 |
| elbflr     | 26, 52     | Motor score C5  | Right | 0,1,2,3,4,5 |
| wrextr     | 26, 52     | Motor score C6  | Right | 0,1,2,3,4,5 |
| elbexr     | 26, 52     | Motor score C7  | Right | 0,1,2,3,4,5 |
| finflr     | 26, 52     | Motor score C8  | Right | 0,1,2,3,4,5 |
| finabr     | 26, 52     | Motor score T1  | Right | 0,1,2,3,4,5 |
| hipflr     | 26, 52     | Motor score L2  | Right | 0,1,2,3,4,5 |
| kneetr     | 26, 52     | Motor score L3  | Right | 0,1,2,3,4,5 |
| ankdor     | 26, 52     | Motor score L4  | Right | 0,1,2,3,4,5 |
| gretor     | 26, 52     | Motor score L5  | Right | 0,1,2,3,4,5 |
| ankplr     | 26, 52     | Motor score S1  | Right | 0,1,2,3,4,5 |



### File 3: test_features.csv

All features, including ISNCSCI assessments and AIS grades, at 1 week after injury. Each row corresponds to a specific patient ID, columns are features.

```{r , echo=FALSE,message=FALSE}
train_outcomes_MS <- read_csv("G:/My Drive/Thesis PhD Boris Polanco/ASIA_CHALLENGE_SCI/ASIAChallenge_ShareFile/test_features.csv",col_names = TRUE)

# Load required libraries
library(dplyr)

# Check structure of dataset
str(train_outcomes_MS)

# Convert all numeric variables to factors (if they are categorical codes)
train_outcomes_MS <- train_outcomes_MS %>%
  mutate(across(where(is.numeric), as.factor))

# Get summary statistics for categorical variables
summary(train_outcomes_MS)

# Get unique value counts per categorical variable
categorical_summary <- train_outcomes_MS %>%
  summarise(across(everything(), ~ n_distinct(.)))

# Count missing values per variable
missing_values <- colSums(is.na(train_outcomes_MS))

# Combine summary into a dataframe
summary_df <- data.frame(
  Variable = names(train_outcomes_MS),
  Unique_Values = unlist(categorical_summary),
  Missing_Values = missing_values
)

# Print summary
print(summary_df)


```

### File 4: test_outcomes_MS_template.csv

All features, including ISNCSCI assessments and AIS grades, at 1 week after injury. Each row corresponds to a specific patient ID, columns are features.

```{r , echo=FALSE,message=FALSE}
test_outcomes_MS <- read_csv("G:/My Drive/Thesis PhD Boris Polanco/ASIA_CHALLENGE_SCI/ASIAChallenge_ShareFile/test_outcomes_MS_template.csv",col_names = TRUE)

```
